---
title: "Differential Expression for multiple contrasts"
author: "NNNNN LLLLL"
date: "`r Sys.Date()`"
output:
    html_document:
        code_folding: hide
        df_print: paged
        highlight: pygments
        number_sections: false
        self_contained: true
        theme: paper
        toc: true
        toc_float:
            collapsed: true
            smooth_scroll: false
params:
    seFile: "data/rna_se.rda"
    design: "~time + condition"
    contrast: "condition"
    conditions: !r c("condition", "time") # first one used in some plots
    alpha: 0.05
    lfc: 0
    outputDir: "."
    cache_dir: "../cache"
---

```{r knitr-setup}
# Set seed for reproducibility
set.seed(1454944673)
library(knitr)
library(ggplot2)

opts_chunk[["set"]](
    autodep = TRUE,
    bootstrap.show.code = FALSE,
    cache = TRUE,
    cache.lazy = TRUE,
    cache.path = params$cache_dir,
    dev = c("png", "pdf"),
    error = TRUE,
    fig.height = 10,
    fig.retina = 2,
    fig.width = 10,
    highlight = TRUE,
    message = FALSE,
    prompt = TRUE,
    # formatR required for tidy code
    tidy = TRUE,
    warning = FALSE)

theme_set(
    theme_light(base_size = 14))
theme_update(
    legend.justification = "center",
    legend.position = "bottom")

```

```{r setup, message=FALSE}
library(DESeq2)
library(SummarizedExperiment)
library(DEGreport)
library(pheatmap)
library(dplyr)
library(readr)
# Load bcbioRNASeq object
load(params$seFile)
bcb <-se
colData(bcb)[["time"]] = gsub("-[0-9]$", "", colData(se)[["sample"]])
colData(bcb)[["condition"]] = as.factor(colData(bcb)[["condition"]])
lfc = params$lfc
alpha = params$alpha
# Directory paths
outputDir <- params$outputDir
dataDir <- dirname(params$seFile)
countsDir <- file.path(outputDir, "results", "counts")
dir.create(countsDir, showWarnings = FALSE, recursive = TRUE)
deDir <- file.path(outputDir, "results", "differential_expression")
dir.create(deDir, showWarnings = FALSE, recursive = TRUE)
```


```{r dds, results="hide", eval=!file.exists(file.path(dataDir, "dds.rda"))}
# help("design", "DESeq2")
dds <- DESeqDataSetFromMatrix(
    countData = assay(bcb),
    colData = colData(bcb),
    design = formula(params$design)) %>%
    DESeq()
rld <- varianceStabilizingTransformation(dds)
save(dds, rld, file = file.path(dataDir, "dds.rda"))
```

# PCA

Let's take a look at the PCA analysis.

```{r general-pca, fig.width=6, fig.height=6}
degPCA(assays(bcb)[["vst"]], colData(bcb), condition = params$condition[1]) +
    ggrepel::geom_text_repel(aes(label=sample))
```


# Results

```{r res, eval=!file.exists(file.path(dataDir, "comparisons.rda"))}
# ?degComps: Read how to get the different contrasts
# for coefficients from one column: degComps(dds, combs = column_name)
# for all possible paris: degComps(dds, combs = column_name, pairs = TRUE)
# for specific contrasts: degComps(dds, contrasts = list(c(column_name, group1, group2),
#                                                        c(column_name, group1, group3)))

comparisons = degComps(dds, combs = params$contrast, type = "ashr")

save(comparisons, file = file.path(dataDir, "comparisons.rda"))
```

```{r load, results="hide"}
lapply(file.path(dataDir, c("dds.rda", "comparisons.rda")), load, environment()) %>% invisible()
```

We performed the analysis using a BH adjusted *P* value cutoff of `r params$alpha` and a log fold-change (LFC) ratio cutoff of `r params$lfc`.

## Alpha level (FDR) cutoffs {.tabset}

Let's take a look at the number of genes we get with different false discovery rate (FDR) cutoffs. These tests subset *P* values that have been multiple test corrected using the Benjamini Hochberg (BH) method [@Benjamini:1995ws].

```{r alpha_summary, results="asis"}
lapply(names(comparisons), function(x){
    cat("### ", x, "\n")
    degSummary(comparisons[[x]], kable = TRUE) %>%  show
    cat("\n\n")
}) %>%  invisible()
```


# Plots

## Mean average (MA) {.tabset}

An MA plot compares transformed counts on `M` (log ratio) and `A` (mean average) scales [@Yang:2002ty].
Blue arrows represent a correction of the log2 Fold Change values due to variability. Normally this happens when the gene shows high variation and the
log2FC is not accurate, here the model tries to estimate them. See this paper for more information:

Love, M.I., Huber, W., Anders, S. (2014) Moderated estimation of fold change and dispersion for RNA-seq data with DESeq2. Genome Biology, 15:550. https://doi.org/10.1186/s13059-014-0550-8

```{r plot_ma, results="asis"}
lapply(names(comparisons), function(x){
    cat("### ", x, "\n")
    print(degMA(comparisons[[x]]))
    cat("\n\n")
}) %>%  invisible()
```


## Volcano {.tabset}

A volcano plot compares significance (BH-adjusted *P* value) against fold change (log2) [@Cui:2003kh; @Li:2014fv]. Genes in the green box with text labels have an adjusted *P* value are likely to be the top candidate genes of interest.

```{r plot_volcano, results="asis"}
lapply(names(comparisons), function(x){
    cat("### ", x, "\n")
    print(degVolcano(comparisons[[x]]))
    cat("\n\n")
}) %>%  invisible()
```


## Heatmap {.tabset}

This plot shows only differentially expressed genes on a per-sample basis. We have scaled the data by row and used the `ward.D2` method for clustering [@WardJr:1963eu].

```{r plot_heatmap, results="asis"}
lapply(names(comparisons), function(x){
    cat("### ", x, "\n")
    s = significants(comparisons[[x]])
    if (length(s) < 2){
        cat("Less than two genes, no possible to plot")
    }else{
        p=pheatmap(assays(bcb)[["vst"]][s,], scale = "row",
                 annotation_col = as.data.frame(colData(bcb)[,params$conditions, drop=F]),
                 show_rownames = FALSE, show_colnames = FALSE, clustering_method = "ward.D2",
                 clustering_distance_cols = "correlation")
        print(p)
    }
    cat("\n\n")
}) %>%  invisible()

```

## PCA {.tabset}

Principal component analysis is a technique to reduce the dimensionality of the data to allow visualization in two dimensions [PCA][]. It takes all the gene abundances for the samples and
creates a series of principal components (PCs) to explain the
variability in the data. We normally plot the first two PCs for
simplicity.

```{r plot_pca, results="asis"}
lapply(names(comparisons), function(x){
    cat("### ", x, "\n")
    s = significants(comparisons[[x]])
    if (length(s) < 2){
        cat("Less than two genes, no possible to plot")
    }else{
        degPCA(assays(bcb)[["vst"]][s,],
               colData(bcb), condition = params$conditions[1]) %>% print
        
        cat("\n\n")
    }
}) %>%  invisible()
```

## Gene Expression Patterns

In general, it is useful to cluster the significant genes together in similar patterns across samples. `degPatterns` uses standard expression correlation technique to generate a similarity matrix that can be clustered hierarchically and then split into groups of genes that follow similar expression patterns [degPtterns][].

We defined significance as genes with abs(log2FC) >  `r cat(lfc)` and FDR < `r cat(alpha)`.

```{r patterns}
# choose the significants genes. 
# In this case the first is used. `sig` should be a character vector with genes.
sig = significants(comparisons, fc = params$lfc, padj = params$alpha)
pattern = degPatterns(assays(bcb)[["vst"]][sig,],
                      colData(bcb), time = params$conditions[1])
```

## Top genes {.tabset}

```{r results_tables, results="asis"}
resTbl = lapply(names(comparisons), function(x){
    cat("### ", x, "\n")
    p = degPlot(bcb, genes = significants(comparisons[[x]])[1:9],
            xs = params$conditions[1],
            slot = "vst", log2 = FALSE,
            ann = c("gene_id", "gene_name"))
    print(p)
    cat("\n\n")
    res = deg(comparisons[[x]], tidy = "tibble") %>% 
        left_join(rowData(bcb) %>% as.data.frame, by = c("gene" = "gene_id")) %>% 
        left_join(pattern[["df"]], by = c("gene" = "genes"))
   dir.create(file.path(deDir, x), showWarnings = FALSE, recursive = TRUE)
   write_csv(res, file.path(deDir, x, paste0(x, ".csv.gz")))
    res
})
names(resTbl) = names(comparisons)
```


## Top tables {.tabset}

Top 10 genes are shown order by False Discovery Rate. Genes below 0.05 are 
considered significant.

```{r top_tables, results="asis"}
lapply(names(resTbl), function(x){
    cat("### ", x, "\n")
    head(resTbl[[x]], 10) %>% kable %>% show
    cat("\n\n")
}) %>%  invisible()
```


# File downloads

The results are saved as gzip-compressed comma separated values (CSV). Gzip compression is natively supported on [macOS][] and Linux-based operating systems. If you're running Windows, we recommend installing [7-Zip][]. CSV files can be opened in [Excel][] or [RStudio][].


## Count matrices

Tables are under `r file.path(countsDir)` folder:

- [`normalizedCounts.csv.gz`](`r file.path(countsDir, "normalizedCounts.csv.gz")`): Use to evaluate individual genes and/or generate plots. These counts are normalized for the variation in sequencing depth across samples.
- [`tpm.csv.gz`](`r file.path(countsDir, "tpm.csv.gz")`): Transcripts per million, scaled by length and also suitable for plotting.
- [`rawCounts.csv.gz`](`r file.path(countsDir, "rawCounts.csv.gz")`): Only use to perform a new differential expression analysis. These counts will vary across samples due to differences in sequencing depth, and have not been normalized. Do not use this file for plotting genes.

## Differential expression tables

Tables are under `r file.path(deDir)` folder:

DEG tables are sorted by BH-adjusted P value, and contain the following columns:

- `ensgene`: [Ensembl][] gene identifier.
- `baseMean`: Mean of the normalized counts per gene for all samples.
- `log2FoldChange`: log2 fold change.
- `lfcSE`: log2 standard error.
- `stat`: Wald statistic.
- `pvalue`: Walt test *P* value.
- `padj`: BH adjusted Wald test *P* value (corrected for multiple comparisons; aka FDR).
- `externalGeneName`: [Ensembl][] name (a.k.a. symbol).
- `description`: [Ensembl][] description.
- `geneBiotype`: [Ensembl][] biotype (e.g. `protein_coding`).


# Methods

RNA-seq counts were generated by [bcbio][] and [bcbioRNASeq][] using [salmon][] [@salmon]. Counts were imported into [R][] using [tximport][] [@tximport] and [DESeq2] [@DESeq2]. Gene annotations were obtained from [Ensembl][]. Plots were generated by [ggplot2][] [@ggplot2]. Heatmaps were generated by [pheatmap][] [@pheatmap].



# R session information {.tabset}

```{r session_info}
devtools::session_info()
print(params)
```



[bcbio]: https://github.com/chapmanb/bcbio-nextgen
[bcbioRNASeq]: http://bioinformatics.sph.harvard.edu/bcbioRNASeq
[Bioconductor]: https://bioconductor.org
[DESeq2]: https://bioconductor.org/packages/release/bioc/html/DESeq2.html
[Ensembl]: http://useast.ensembl.org
[Excel]: https://products.office.com/en-us/excel
[ggplot2]: http://ggplot2.org
[macOS]: https://www.apple.com/macos
[pheatmap]: https://cran.r-project.org/web/packages/pheatmap/index.html
[R]: https://www.r-project.org
[RStudio]: https://www.rstudio.com
[salmon]: https://combine-lab.github.io/salmon
[tximport]: https://bioconductor.org/packages/release/bioc/html/tximport.html
[7-Zip]: http://www.7-zip.org
[PCA]: https://en.wikipedia.org/wiki/Principal_component_analysis
[degPattern]: https://lpantano.github.io/DEGpattern
